\chapter{\januso{}}
\label{chap:janus1}

In this chapter we will extend the \janusz{} language with new
concepts, forming the language \januso{}. The \januso{} language adds
several extra inference rules to the \janusz{} language in order to
make it more in par with the full complete language
specification. Concretely, we add:
\begin{itemize}
\item 32-bit unsigned integers with wrap-around arithmetic $(\mod
  2^{32})$.
\item Procedure calls
\end{itemize}
\januso{} only lacks arrays and loops from the full Janus language.

We first introduce 32-bit integers in \coq{} and then use the
formalization of these in the development of \januso{}.

\section{32-bit Integers}
\label{sec:32-bit-integers}

The integers chosen in JANUS are as usual machine words on a 32-bit
machine. Most traditional computers have an upper limit to what
numbers it can represent. A 32-bit machine has $32$ bits at its
disposal so it can represent $2^{32}$ different values. A 64-bit
machine has $2^{64}$ different values and so on. For the case where
there is no sign, it is obvious to map the representation into $0,
\dotsc, 2^{32}-1$ and do calculation modulo $2^{32}$. Coincidentally,
this is how most modern computers work anyway for unsigned arithmetic.

JANUS chooses this representation as well and models a 32-bit machine
in the process. For \januso{}, we would like to model the same thing.

From the perspective of mathematics, 32-bit values means we are
working in the ring $\mathbb{Z}/2^{32}\mathbb{Z}$
\cite{jensen:klassisk}.  We could have chosen other
rings as the base for our numbers in JANUS.

Let us recap exactly what is meant by a ring
\begin{defn}
  A ring is a set $R$ equipped with two binary compositions $+\colon R
  \times R \to R$ and $\cdot \colon R \times R \to R$. The $+$
  composition forms an abelian group and the $\cdot$ composition forms
  a monoid. Furthermore, the ``multiplication'' distributes over the
  addition by the following rules:
  \begin{align*}
    a \cdot (b + c) & = (a \cdot b) + (a \cdot c)\\
    (a + b) \cdot c & = (a \cdot c) + (b \cdot c)
  \end{align*}
\end{defn}

One particular ring is well-known and a good candidate for a number
base; the usual ring of integers $\mathbb{Z}$. Implicitly,
this was the choice for \janusz{}. From a formal viewpoint, this ring
is bliss to work with. On the other hand, it means that the formal
machine have arbitrarily sized integers, which may be a bit
unrealistic. Hence, we chose numbers modulo $2^{32}$ as we can easily
formalize 64-bit numbers as well if needed by changing a constant. We
will call these integers $W^{32}$.

\begin{rem}
  Note that the ``multiplication'' need only form a monoid and not a
  group in general. We are thus not sure of a multiplicative inverse,
  which is somewhat bad. Remedying this could be to use finite
  Galois-fields, of order $p^n$ where $p$ is a prime and $n$ is an
  integer. For the case $p = p^1$ it is simple since it is defined by
  the groups $\ZZ / p\ZZ$ (for addition) and $(\ZZ / p\ZZ)^{*}$ (for
  multiplication). For the case $p^n$ for $n > 1$ you need to use
  splitting fields and more heavyweight algebra to define them. This
  is turn makes them rather bad candidates as a number base. Perhaps
  except for the base $2^n$ which can be defined as polynomials of
  bits. In any case, Galois-fields are not easily
  machine-implementible in hardware, so we choose not the pursue this
  idea further.
\end{rem}

\subsection{\coq{} Implementation}

The \coq{} implementation of 32-bit numbers owes everything to Xavier
Leroy et.al.\cite{leroy:compcert}. The work is taken from the
\textsc{CompCert} project, which introduces them in the course of
formalizing a complete C to PowerPC compiler. The reason for taking
the work rather than building it from scratch is due to the sheer
complexity of getting it right. The formalization uses standard
algebraic properties to systematically build the necessary knowledge
and properties in \coq{}.

We did some simplification to the full specification as we only needed
parts of it. We also added a couple of new properties to the
development which is needed for proving \januso{}.

We can't resist the temptation to describe the implementation however. A
32-bit number is a pair: An integer $i$ and a proof that $0 \leq i <
2^{32}$. The main theorem is:
\begin{lem}
  Let $x$ be in integer. For any such $x$ we have
  \begin{equation*}
    0 \leq (x \;mod\; 2^{32}) < 2^{32}
  \end{equation*}
\end{lem}
\begin{proof}
  From the \coq{} Development:
\begin{verbatim}
Lemma mod_in_range:
  forall x, 0 <= Zmod x modulus < modulus.
  intros.
  exact (Z_mod_lt x modulus (two_power_nat_pos word_size)).
Qed.
\end{verbatim}
  This states it is exactly given by the common result
  \texttt{Z\_mod\_lt} from the \coq{} standard library.
\end{proof}
The theorem allows us to take any integer and represent it as a 32-bit
number as long as we take the integer $(\mod 2^{32})$. Thus we have a
straightforward embed/project pair of functions between normal
integers and 32 bit integers.

For example, here is the definition of addition:
\begin{verbatim}
Definition add (x y: w32) : w32 :=
  repr (unsigned x + unsigned y).
\end{verbatim}
The function \texttt{unsigned} projects $x$ and $y$ to integers. Then
we integer-add the numbers and re-embed them with the \texttt{repr}
function. Note that the embedding has the correct behaviour: it wraps
around.

Bit-wise operations in the representation is implemented by creating
indicator functions from numbers up to a certain size in bits. Ie. a
number is converted into a function $f \colon \mathbb{Z} \to \{0,
1\}$, where $f(n) = 1$ indicates $n$-th bit is set and $f(n) = 0$ that
it is not. Straightforward combinations, projections, and embeddings
of these functions builds up the bit-wise functions.

The properties of the numbers are simply a direct (rather impressive)
work of standard algebra. The theorems can be found in any
introductory math-book on algebra, for instance
\cite{thorup:algebra}. First, we have
\begin{defn}
  Two integers $x, y$ are in the same congruence class modulo $m$ iff
  there exists an integer $k$ such that
  \begin{equation*}
    x = k \cdot m + y
  \end{equation*}
\end{defn}
With this definition Xavier and co. then proceeds to build up a
complete library of known algebraic properties for congruence classes
like the one above. As an example we present the addition of
congruence classes:
\begin{verbatim}
  Lemma eqmod_add:
    forall a b c d, eqmod a b -> eqmod c d
      -> eqmod (a + c) (b + d).
  Proof.
    intros a b c d [k1 EQ1] [k2 EQ2]; red.
    subst a; subst c. exists (k1 + k2). ring.
  Qed.
\end{verbatim}
If $a, b$ is a congruence class pair and $c, d$ also is, then the pair
$a+c, b+d$ is an equivalence pair (for a fixed modulo). The proof proceeds by the
well-known method that there must exist a $k_1$ by the first pair such
that $a = k_1 \cdot m + b$. Likewise $c = k_2 \cdot m + d$. Adding
these 2 equations and rearranging yields $a + c = (k_1 + k_2) \cdot m
 + (b + d)$ which is the form one wants. Most of the proofs reflect
usual mathematical methods like this.

\paragraph{Additions to the CompCert integers}

We added some additional properties to the \textsc{CompCert}
implementation of 32-bit integers. All the proofs of these are written
in \coq{} by the use of already given properties by Leroy
et.al. We extend their definitions by:
\command{xor[\;xor\;]}
\begin{lem}
  For any $x \in W^{32}$ we have
  \begin{equation*}
    x \<xor> x = 0
  \end{equation*}
\end{lem}
\begin{lem}
  For 32-bit integers $x,y,z$, additions can be discharged on the
  left:
  \begin{equation*}
    x + y = x + z \implies y = z
  \end{equation*}
\end{lem}
\begin{lem}
  For 32-bit integers $x, y, z$, subtraction can be discharged on the
  right:
  \begin{equation*}
    x - z = y - z \implies x = y
  \end{equation*}
\end{lem}
\begin{lem}
  For 32-bit integers $x, y, z$, xor can be discharged on left:
  \begin{equation*}
    x \<xor> z = y \<xor> z \implies x = y
  \end{equation*}
\end{lem}

\section{Augmenting expressions}

\fixme{Write that we changed to 32-bit}
To the expressions of \janusz{}, we make the following additions:
\begin{align*}
  e ::= & \dotsc \bor e / e \bor e \mod e \bor e = e \bor e \neq e \\
        \bor & e \land e \bor e \lor e \bor e < e
\end{align*}

The additions to the expression language are as in \cite{yokoyama.axelsen.ea:principles};
we add arithmetic operators of division and modulo to 32-bit
numbers. We add a couple of relational operators, namely equality,
non-equality, logical and, logical or and a less-than operator.

With these additions, we almost have all constructs from the full
JANUS language. It is our belief that the remaining operators can be
added without any problems. Specifically, our 32-bit integer
representation provisions for bit-wise operators.

The semantics are given by a denotational semantics in the spirit of
the denotational semantics in \ref{exp:denot-semantics}. We omit them
here for brevity and refer the interested reader to the \coq{}
development.

\section{Augmenting statements}

To the statement language we add the rules of \texttt{call},
\texttt{uncall} and xor ($\hat{=}$).
\begin{equation*}
  s ::= \dotsc \bor \<call> p \bor \<uncall> p \bor x \; \hat{=} \; e
\end{equation*}
where $p$ designates a procedure. We will see how we encode this
procedure shortly.

The semantical judgement forms of statements are altered to a form
$\rho |= \angel{\sigma, s} -> \sigma'$, where $\rho$ is a map from
$\ZZ \to Stm$. That is, we represent a procedure by an integer and its
value in the set of procedure definitions is a statement. We note that
all existing rules in the operational semantics just has to pass
$\rho$ on via congruence. Hence, we do not bring these rules here. The
rule of calling a procedure is
\begin{equation*}
  \inference[Call]{\rho |= \angel{\sigma, \rho(p)} -> \sigma'}
  {\rho |= \angel{\sigma, \<call> p} -> \sigma'}
\end{equation*}
Informally, the rule states that a procedure call looks up the
procedure in the context and then executes the procedure. Uncalling a
procedure, running the procedure in inverse, is defined as
\begin{equation*}
  \inference[Uncall]{\rho |= \angel{\sigma', \rho(p)} -> \sigma}
  {\rho |= \angel{\sigma, \<uncall> p} -> \sigma'}
\end{equation*}
In this rule, we utilize the full power of operational-semantics given
as a Prolog-style rule. Rather than relying on the inversion operator
we will define shortly, we alter the order procedure will run. This is
the main reason we choose an operational semantics rather than a
denotational one as the latter would have a harder time representing
this specific rule.

For the xor-operation, we use a rule which reflects the increment and
decrement operator:
\begin{equation*}
  \inference[Xor]{(\sigma \setminus x) |- e => \lift{k} \quad \sigma(x) =
    \lift{k'} \quad k \oplus_{32} k' = n}{\angel{\sigma, x \;\hat{=}\; e} -> \sigma[x \mapsto n]}
\end{equation*}
where $\oplus_{32}$ is the bit-wise xor-operation.

\section{Encoding expressions and statements}

The encoding of expressions and statements follows the same style as
we used earlier. The expression evaluation is extended with the new
constructors and utilizes our encoding of 32-bit words. We take the
existing encoding and gradually adapt it towards \januso{}.

Statements are implemented by first introducing a mapping $\ZZ \to
Stm$ which we then use for the $\rho$ in the above judgement
forms. Then the Inductive definition of statements are altered to take
this $\rho$ as and additional dependent type.

With these encodings, we can capture all of \januso{} and we can turn
our focus towards the properties of the language.

\section{Determinism properties of \januso{}}

Converting the proofs from \janusz{} to \januso{} has one main
complication: we went from $\ZZ$ to $W^{32}$ as our number type. When
\coq{} works with integers, and we only used Presburger arithmetic we
can solve such terms by the use of the \texttt{omega} tactic in
\coq{}. For our 32-bit integers, no such thing is present however, so
we need to do the proofs by hand.

The detail at which one has to carry out such proofs are almost
excruciatingly precise. We must, as an example, explicitly use the
commutative rule when we want to rewrite $x +_{32} y$ into $y +_{32}
x$.

\fixme{This section needs to be rewritten}
However, the formalization in \coq{} is slightly different to work
around a problem with mutually defined theorems. This means we cannot
utilize the theorem in further development as it is given in the
development.
\begin{thm}
\label{thm:j1-fwd-det}
  \januso{} is forward deterministic, ie:
  \begin{equation*}
    \forall \sigma, \sigma', \sigma'' \in \Sigma, s \in Stm \colon
    \rho |= \angel{\sigma, s} -> \sigma' => \rho |= \angel{\sigma, s} -> \sigma'' => \sigma' = \sigma''
  \end{equation*}
\end{thm}
\begin{proof}
  In \coq{}
\end{proof}
and
\begin{thm}
\label{thm:j1-bwd-det}
  \januso{} is backward deterministic, ie:
  \begin{equation*}
    \forall \sigma, \sigma', \sigma'' \in \Sigma, s \in Stm \colon
    \rho |= \angel{\sigma', s} -> \sigma \implies \rho |= \angel{\sigma'', s} -> \sigma \implies \sigma' = \sigma''
  \end{equation*}
\end{thm}
\begin{proof}
  Formalized in \coq{}
\end{proof}

\section{Inversion and its properties}

Inverting \januso{} works exactly like inverting \janusz{}. We only
need to add a couple of new rules for $\<call>$ and $\<uncall>$ and
the $\hat{=}$ operation. The two call types are each others inverses
and the $\<xor>$ operation is self inverse:
\begin{align*}
    \mathcal{I}(x \;\hat{=}\; e)& = x \;\hat{=}\; e\\
    \mathcal{I}(\<call> p)& = \<uncall> p\\
    \mathcal{I}(\<uncall> p)& = \<call> p
\end{align*}

With these rules, we can again prove the inversion sound by the
following theorem:
\begin{thm}
  Let $\rho$ designate an arbitrary \januso{} program, $s$ an
  arbitrary \januso{} statement and let $\sigma$ and $\sigma'$ be
  arbitrary stores. The following is then true:
  \begin{equation*}
    \rho |= \angel{\sigma, s} -> \sigma' \iff \rho |= \angel{\sigma', \mathcal{I}(s)} -> \sigma
  \end{equation*}
\end{thm}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: "master"
%%% End:
