\documentclass[a4paper, oneside, 10pt, draft]{memoir}
\chapterstyle{culver}
\usepackage{fixme}
\usepackage{coqdoc}
\usepackage[english]{babel}
\usepackage[osf,sc]{mathpazo}
\linespread{1.05}
%\usepackage[utopia]{mathdesign}
\input{../../Preamble/preamble_memoir.tex}

\usepackage{semantic}
\author{Jesper Louis
  Andersen\\jesper.louis.andersen@gmail.com\\140280-2029}
\title{PIRC - Exam}
\date{\today}

\renewcommand*{\titleM}{\begingroup% Misericords, T&H p 153
  \drop = 0.08\textheight
  \centering
  {\Huge\bfseries Formalizing JANUS}\\[\baselineskip]
  {\scshape In a proof assistant}\\[\baselineskip]
  {\scshape by}\\[\baselineskip]
  {\large\scshape Jesper Louis Andersen\\jesper.louis.andersen@gmail.com}\par
  \endgroup}

\bibliographystyle{plain}
\hyphenation{in-di-ca-tor func-tions}

\begin{document}
\newcommand{\janusz}{$\mathrm{JANUS}_0$}
\newcommand{\januso}{$\mathrm{JANUS}_1$}
\newcommand{\lift}[1]{\lfloor #1 \rfloor}
\newcommand{\coq}{{\scshape Coq}}
\newcommand{\twelf}{{\scshape Twelf}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\titleM
\begin{abstract}
  In this project, we provide a state-of-the-art formalization on the
  reversible language JANUS. We formalize the syntax, the semantics
  and we provide verification of main reversibility proofs in the
  proof assistant \coq{} amounting to over 1500 lines of vernacular
  definitions. The formalization covers large parts of JANUS and
  points to places that are impossible to prove with the currently
  given semantics. Furthermore we provide improvements on the existing
  semantics by formalizing properties that were stated informally in
  the existing litterature.
\end{abstract}
\listoffixmes
\tableofcontents
\fixme{Motivating introduction}
\fixme{Overview where censor can view all of the report}
\fixme{Expectations}
\fixme{Think about readability of the report}
\fixme{Examples of carrying out proofs in \coq{}}
\fixme{Ispell from here once everything is finished}
\input{intro}
\input{cocext}
\fixme{Ispell from here once we have it all.}
\fixme{Read through J0, correcting.}
\input{janus0}
\fixme{Read through J1, correcting.}
\input{janus1}
\fixme{Read through J, correcting.}
\input{fulljanus}
\fixme{Read through FW, correcting.}
\fixme{Change further work to related work.}
\input{furtherwork}
\fixme{Read through Conc., correcting}
\fixme{Write about stacks and recursive languages as a solution}
\fixme{Write about a forth-variant as a solution}
\input{conclusion}
\bibliography{biblio}
\appendix
\input{source}
\end{document}